High geomagnetic field intensity recorded by anorthosite xenoliths requires a strongly powered late Mesoproterozoic geodynamo

### 2022-02-04 PNAS_manuscript/PNAS_manuscript.tex

\affil[a]{Department of Earth and Planetary Science, University of California, Berkeley, CA, 94720}
\affil[b]{Geology, Minerals, Energy, and Geophysics Science Center, U.S. Geological Survey, Moffett Field, CA, 94025}
\affil[c]{Department of Earth and Planetary Sciences, Harvard University, Cambridge, MA, 02138}

@@

Despite the excellent preservation of the rocks, non-ideal paleointensity behaviors have challenged the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a}. The most trusted type of paleointensity estimate is that obtained through experiments in which the primary natural thermal remanent magnetization is progressively replaced by a laboratory magnetization that is imparted in a known field with internal consistency checks (such as in IZZI-style Thellier experiments;  \citealp{Yu2004a}). In such Thellier paleointensity experiments, one typical departure from ideal behavior is sagging or double-slopes as visualized in plots that show thermal remanent magnetization (TRM) acquired versus natural remanent magnetization (NRM) lost. For such data, distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope. Typically, such non-ideal behavior would result in a higher paleointensity estimate from the steeper-sloped low-temperature portion and a lower paleointensity estimate from the high-temperature portion. For example, in data from the Midcontinent Rift, ref.  \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Such non-ideal results were rejected by \citealp{Sprain2018a} who applied stricter paleointensity selection criteria, but as a result had few accepted sites.

@@

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=13 cm]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic images (A,B,C,D) and magnetic maps (E,F) of anorthosite from the Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column;  \citealp{Zhang2021b}) and a distinct anorthosite from the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given the complexities associated with more pronounced fabrics. Cross-polarized petrographic images of the Beaver River anorthosite (A,C) reveal plagioclase with a granoblastic texture with crystals that are largely free of large opaque inclusions. In contrast, plagioclase crystals in the Duluth Complex anorthosite have euhedral, interlocking crystals with an igneous foliation (B) and the plagioclase crystals contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [001] axis of the plagioclase. Magnetic field maps acquired with a quantum diamond microscope (QDM) on samples to which an IRM has been applied show that the relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite in (E) while large oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the overlapping magnetic fields above the oxides in (F). Magnetic field maps of (E) and (F) share the same spatial and magnetic scale. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

@@

The dominantly monomineralic anorthosite xenoliths within the Beaver River diabase often have granoblastic texture characterized by equigranular crystals with weakly developed petrofabrics (Fig. \ref{fig:Petro_QDM}A). In addition, the plagioclase crystals of the anorthosite xenoliths are typically free of large oxide inclusions (Fig. \ref{fig:Petro_QDM}C). Magnetic imaging using a quantum diamond microscope (QDM) shows that the dominant magnetic remanence carriers often exist within, and sometimes interstitial to, plagioclase crystals (Fig. \ref{fig:Petro_QDM}E).

@@

Coercivity spectra from backfield experiments further support the interpretation that rocks that pass or fail the selection criteria have distinct magnetic mineralogy (Fig. \ref{fig:coercivity}). A compilation of median destructive field (MDF) values in Figure. \ref{fig:coercivity} shows that specimens from anorthosites that pass paleointensity selection criteria have distinctly higher average MDF values than other anorthosite and diabase specimens. Single-component fits for coercivity spectra \cite{Maxbauer2016a} show that anorthosites that yielded successful paleointensity results can have magnetic grain populations with peak coercivity around 80 mT (Fig. \ref{fig:coercivity}). In contrast, other anorthosite and diabase tend to have lower peak coercivities ($\sim$30 mT). This result is consistent with an interpretation that a population of magnetic grains with more multidomain-like behavior is responsible for the non-ideal paleointensity behaviors during experiments on such specimens.

@@

\subsection*{TRM anisotropy and nonlinear acquisition check}
Significant remanence anisotropy has been documented to exist within certain anorthositic rocks that form in layered intrusive complexes \cite{Selkin2000a, Feinberg2006a}. Strong remanence anisotropy associated with the igneous foliation developed within anorthosite from the Stillwater Complex has been shown to lead to significant overestimation or underestimation of paleointensity values depending on the relative orientations between the fabrics and an applied magnetic field \cite{Selkin2000a}. To assess whether our paleointensity estimates are biased by remanence anisotropy, we calculated the gamma statistic, which is the angular difference between the last pTRM step of paleointensity experiment and the applied field direction. The results show that the anorthosite specimens used for paleointensity experiment have low gamma values ranging from 0.9\textdegree$\;$ to 11.9\textdegree, with a median value of 4.2\textdegree$\;$(Table S1). These gamma values are similar to those of Midcontinent Rift volcanics \cite{Sprain2018a}. Therefore, the Beaver River anorthosite xenolith bulk samples do not have significant remanence anisotropy. Paleodirectional data from our anorthosite xenoliths further support that they have minimal remanence anisotropy as their site mean directions closely match those of the Beaver River diabase hosts without deviating due to a fabric \cite{Zhang2021b}. In addition, no resolvable magnetic fabric was observed in the anorthosite xenoliths through petrographic and magnetic imaging (Fig. \ref{fig:Petro_QDM}).

@@

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{PINT_compilation.pdf}
\centering
\caption{\footnotesize{Compilation of calculated virtual (axial) dipole moment values from the PINT database (PINT v8.0.0; \url{http://www.pintdb.org/};  \citealp{Bono2021a}), including all Phanerozoic VDM and VADM records with Q$_{PI}$ values $>$3 and additional Neoproterozoic data from refs. \citealp{Lloyd2021a, Lloyd2021b, Thallner2021a, Thallner2021b}. Paleointensity estimates from \cite{Pesonen1983a} and \cite{Kulakov2013a} are not included in the compilation due to the specimen-level double-slope behavior as discussed in the text. Overall the anorthosite xenoliths from this study record a high Mesoproterozoic field exceeding the value projected by the second order polynomial curve from \cite{Bono2019a} which is based on an interpretation of there being a monotonic decay of the geodynamo through the Proterozoic. The highest site-mean virtual dipole moment of the anorthosites are similar to the 99.5 percentile of the selected Phanerozoic data. }}
\label{fig:PINT_compilation}
\end{figure*}

@@

%The outline for this section is::
%- Discussion of mineral physics constraints (or lack thereof)
%- Predictions for paleointensity signal associated with inner core nucleation (or lack thereof)
%- Interpretations of paleomagnetic record (i.e. Bono model of monotonic decrease and then rapid increase;. Cite Biggin et al. 2015) related to inner core nucleation. 
%- Challenges posed to this model based on our data and other low to strong field transitions.

@@

The other data type that can provide insight into the long-term history of the core's thermal regime and geodynamo is paleomagnetic data---both paleodirectional data that indicate the presence of a geomagnetic field and paleointensity data that constrain the field's strength. Inner core nucleation would have increased the power to the geodynamo which has the potential to manifest as an increase in Earth's surface field \cite{Davies2021a}. An approach combining dynamo simulations and theoretical scaling relationships has predicted that progressive decay of the field's dipole moment would be followed by a rapid increase in geomagnetic field intensity soon after the onset of inner core nucleation such that there is a minimum in dipole moment just before inner core nucleation \cite{Davies2021a}. Other scenarios are possible, however, such as the prediction that while power increases associated with inner core nucleation would result in increases in Earth's internal magnetic field strength, the dynamo would also become more deeply seated in the core, diminishing the increase in observed magnetic field at Earth's surface \cite{Aubert2009a, Landeau2017a}. Observational paleomagnetic records hold the potential for testing different model predictions and identifying transitions in ancient field strength \cite{Biggin2015a, Bono2019a}.

@@

The high paleointensity estimates from the 1.1 billion-year-old Midcontinent Rift rocks challenge the hypothesized monotonic decay of the strength of the geomagnetic field throughout the Proterozoic Era. Instead, the well-preserved ca. 1092 Ma anorthosite xenoliths of the Beaver River diabase record a strong geomagnetic field in the late Mesoproterozoic that exceeds the strength of the modern-day field for which crystallization of the inner core is a power source (Fig. \ref{fig:PINT_compilation}). Together with previous records obtained from the ca. 1106 Ma Osler Volcanics of the Midcontinent Rift \cite{Sprain2018a}, these data indicate that appreciable power to Earth's dynamo persisted through at least 14 Myr during the late Mesoproterozoic to maintain a strong surface field (Fig. \ref{fig:PINT_compilation}). In addition to these high geomagnetic fields recorded by Midcontinent Rift rocks, the ca. 755 Ma Mundine Well dikes \cite{Lloyd2021b} also require a stronger geomagnetic field in the Neoproterozoic than would be predicted by a progressive Proterozoic decline (Fig. \ref{fig:PINT_compilation}).

The hypothesis that a weak Ediacaran geomagnetic field is a telltale sign of the lack of an inner core with core nucleation following shortly thereafter may predict that it is the most significant weak to strong field transition in the paleointensity record. However, Fig. \ref{fig:PINT_compilation} shows that transitions from low to high field intensities occurred before, during, and after the Ediacaran. In the Ediacaran record developed to date, there is a two-fold increase in Earth's virtual dipole moment when comparing estimates from the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} to those from ca. 550 Ma volcanics of the Skinner Cove Formation \cite{Thallner2021a} (Fig. \ref{fig:PINT_compilation}). In the late Mesoproterozoic, there is at least a six-fold increase within a period of $\sim$35 Myr from a low average virtual dipole moment of $\sim$13 ZAm$^2$ recorded by the ca. 1140 Ma Abitibi dikes which yielded straight Arai plots \cite{Macouin2003a}, to a high moment of $\sim$70 ZAm$^2$ recorded by the ca. 1106 Ma Osler Volcanics, with even stronger values from ca. 1092 Ma by Beaver River anorthosite xenoliths that record virtual dipole moments up to $\sim$140 ZAm$^2$. While the Abitibi dikes paleointensity estimates do not go to values as low as the Sept-$\hat{I}$les intrusions, this virtual dipole moment increase in the Mesoproterozoic is the largest yet documented in the Precambrian on a 10s of millions of years timescale (Fig. \ref{fig:PINT_compilation}). The tempo and scale of this field intensity transition could match with model-based predictions associated with the onset of inner core nucleation \cite{Davies2021a}. This timing would be broadly consistent with that proposed by \citealp{Biggin2015a}. However, a model prediction of sustained strong field values following inner core nucleation is challenged by data from the ca. 1070 Ma Bangemall Sills which include a sill with a low virtual dipole moment of $\sim$6.4 ZAm$^2$ (Fig. \ref{fig:PINT_compilation};  \citealp{Lloyd2021b}). Following the Ediacaran, there are also low paleointensity estimates from Devonian rocks such as ca. 370 Ma dikes and lavas of the Siberian Viluy Traps that give virtual dipole moment estimates of 4.3 to 14.9 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2019a}). These low values as well as data from the ca. 414 Ma Strathmore lava flows \cite{Hawkins2021a}, the ca. 410-380 Ma lava flows of Siberia and the Kola Peninsula \cite{Shcherbakova2017a}, the ca. 408–393 Ma Buribay volcanics \cite{Shcherbakova2021a} and the ca. 332 Ma Kinghorn volcanics \cite{Hawkins2021a} has led to the proposal of this interval as the ``Mid-Paleozoic Dipole Low'' \cite{Hawkins2021a}. This ``Mid-Paleozoic Dipole Low'' is followed by high paleointensity values such that there is a six-fold increase in the database from virtual dipole moment of $\sim$16 ZAm$^2$ at ca. 332 Ma to $\sim$99 ZAm$^2$ ca. 308 Ma in the late Carboniferous (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2021a}).

>>>>

\affil[a]{Department of Earth and Planetary Science, University of California, Berkeley, CA, 94720}
\affil[b]{U.S. Geological Survey, Moffett Field, CA, 94025}
\affil[c]{Department of Earth and Planetary Sciences, Harvard University, Cambridge, MA, 02138}

@@

Despite the excellent preservation of the rocks, non-ideal paleointensity behaviors have challenged the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a}. The most trusted type of paleointensity estimate is that obtained through experiments in which the primary natural thermal remanent magnetization is progressively replaced by a laboratory magnetization that is imparted in a known field with internal consistency checks (such as in IZZI-style Thellier experiments;  \citealp{Yu2004a}). In such Thellier paleointensity experiments, one typical departure from ideal behavior due to the presence of nonuniformly magnetized grains and alteration during experiments is sagging or double-slopes as visualized in plots that show thermal remanent magnetization (TRM) acquired versus natural remanent magnetization (NRM) lost. For such data, distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope. Typically, such non-ideal behavior would result in a higher paleointensity estimate from the steeper-sloped low-temperature portion and a lower paleointensity estimate from the high-temperature portion. For example, in data from the Midcontinent Rift, ref.  \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Such non-ideal results were rejected by \citealp{Sprain2018a} who applied stricter paleointensity selection criteria, but as a result had few accepted sites.

@@

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=13 cm]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic images (A,B,C,D) and magnetic maps (E,F) of anorthosite from the Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column;  \citealp{Zhang2021b}) and a distinct anorthosite from the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given the complexities associated with more pronounced fabrics. Cross-polarized petrographic images of the Beaver River anorthosite (A,C) reveal plagioclase with a granoblastic texture with crystals that are largely free of large opaque inclusions. In contrast, plagioclase crystals in the Duluth Complex anorthosite have euhedral, interlocking crystals with an igneous foliation (B) and the plagioclase crystals contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [001] axis of the plagioclase. Magnetic field maps acquired with a quantum diamond microscope (QDM) on samples to which an IRM has been applied show that the relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite in (E) while large oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the overlapping magnetic fields above the oxides in (F). Magnetic field maps of (E) and (F) share the same spatial and magnetic scale. B_z denotes magnetic field intensity in the vertical direction with positive values corresponding to the out of page direction. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

@@

The dominantly monomineralic anorthosite xenoliths within the Beaver River diabase often have granoblastic texture characterized by equigranular crystals with weakly developed petrofabrics (Fig. \ref{fig:Petro_QDM}A). In addition, the plagioclase crystals of the anorthosite xenoliths are typically free of large oxide inclusions with high aspect ratios (Fig. \ref{fig:Petro_QDM}C). Magnetic imaging using a quantum diamond microscope (QDM) shows that the dominant magnetic remanence carriers often exist within, and sometimes interstitial to, plagioclase crystals (Fig. \ref{fig:Petro_QDM}E).

@@

Coercivity spectra from backfield experiments further support the interpretation that rocks that pass or fail the selection criteria have distinct magnetic mineralogy (Fig. \ref{fig:coercivity}). A compilation of median destructive field (MDF) values in Figure \ref{fig:coercivity} shows that specimens from anorthosites that pass paleointensity selection criteria have distinctly higher average MDF values than other anorthosite and diabase specimens. Single-component fits for coercivity spectra \cite{Maxbauer2016a} show that anorthosites that yielded successful paleointensity results can have magnetic grain populations with peak coercivity around 80 mT (Fig. \ref{fig:coercivity}). In contrast, other anorthosite and diabase tend to have lower peak coercivities ($\sim$30 mT). This result is consistent with an interpretation that a population of magnetic grains with more multidomain-like behavior is responsible for the non-ideal paleointensity behaviors during experiments on such specimens \cite{Xu2004a}.

@@

\subsection*{TRM anisotropy and nonlinear acquisition check}
Significant remanence anisotropy has been documented to exist within certain anorthositic rocks that form in layered intrusive complexes \cite{Selkin2000a, Feinberg2006a}. Strong remanence anisotropy associated with the igneous foliation developed within anorthosite from the Stillwater Complex has been shown to lead to significant overestimation or underestimation of paleointensity values depending on the relative orientations between the fabrics and an applied magnetic field \cite{Selkin2000a}. To assess whether our paleointensity estimates are biased by remanence anisotropy, we calculated the gamma statistic, which is the angular difference between the last pTRM step of paleointensity experiment and the applied field direction. The results show that the anorthosite specimens used for paleointensity experiment have low gamma values ranging from 0.9\textdegree$\;$ to 11.9\textdegree, with a median value of 4.2\textdegree$\;$(Table S1). Because these anorthosite specimens were collected with various orientations with respect to the outcrops, the angle between the applied lab field direction during paleointensity experiments with respect to any fabrics within the anorthosites specimens are expected to cover a range of angles. In addition, these gamma values are similar to those of Midcontinent Rift volcanics \cite{Sprain2018a}. Therefore, the Beaver River anorthosite xenolith bulk samples do not have significant remanence anisotropy. Paleodirectional data from our anorthosite xenoliths further support that they have minimal remanence anisotropy as their site mean directions closely match those of the Beaver River diabase hosts without deviating due to a fabric \cite{Zhang2021b}. In addition, no resolvable magnetic fabric was observed in the anorthosite xenoliths through petrographic and magnetic imaging (Fig. \ref{fig:Petro_QDM}).

@@

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{PINT_compilation.pdf}
\centering
\caption{\footnotesize{Compilation of calculated virtual (axial) dipole moment values from the PINT database (PINT v8.0.0; \url{http://www.pintdb.org/};  \citealp{Bono2021a}), including all Phanerozoic VDM and VADM records with Q$_{PI}$ values $>$3 and additional Neoproterozoic data from refs. \citealp{Lloyd2021a, Lloyd2021b, Thallner2021a, Thallner2021b}. Paleointensity estimates from \cite{Pesonen1983a} and \cite{Kulakov2013a} are not included in the compilation due to the specimen-level double-slope behavior as discussed in the text. Overall the anorthosite xenoliths from this study record a high Mesoproterozoic field exceeding the value projected by the second order polynomial curve from \cite{Bono2019a} which is based on an interpretation of there being a monotonic decay of the geodynamo through the Proterozoic. The highest site-mean virtual dipole moment of the anorthosites would rank in the top 1$\%$ of those in the database for the Cenozoic Era (the last 66 million years) when there was unequivocally a crystallizing inner core. }}
\label{fig:PINT_compilation}
\end{figure*}

@@

@@

The other data type that can provide insight into the long-term history of the core's thermal regime and geodynamo is paleomagnetic data---both paleodirectional data that indicate the presence of a geomagnetic field and paleointensity data that constrain the field's strength. Inner core nucleation would have increased the power to the geodynamo which has the potential to manifest as an increase in Earth's surface field \cite{Davies2021a}. An approach combining dynamo simulations and theoretical scaling relationships has predicted that progressive decay of the field's dipole moment would be followed by a rapid increase in geomagnetic field intensity soon after the onset of inner core nucleation such that there is a minimum in dipole moment just before inner core nucleation \cite{Davies2021a}. Other scenarios are possible, however, such as the model-based prediction that while power increases associated with inner core nucleation strengthen Earth's internal magnetic field, that the dynamo becomes more deeply seated in the core diminishes the increase in magnetic field strength at Earth's surface \cite{Aubert2009a, Landeau2017a}. Such a shift of the dynamo to a greater depth associated with inner core nucleation led \cite{Aubert2009a} to conclude that the increase in power to the dynamo would be difficult to detect with paleointensity data. Ultimately, further observational paleomagnetic records is key as they hold the potential for testing different model predictions and identifying transitions in ancient field strength \cite{Biggin2015a, Bono2019a}.

@@

The high paleointensity estimates from the 1.1 billion-year-old Midcontinent Rift rocks challenge the hypothesized monotonic decay of the strength of the geomagnetic field throughout the Proterozoic Era (Fig. \ref{fig:PINT_compilation}). The well-preserved ca. 1092 Ma anorthosite xenoliths of the Beaver River diabase record a strong geomagnetic field in the late Mesoproterozoic that exceeds the strength of the modern-day field for which crystallization of the inner core is a power source (Fig. \ref{fig:PINT_compilation}). Together with previous records obtained from the ca. 1106 Ma Osler Volcanics of the Midcontinent Rift \cite{Sprain2018a}, these data indicate that appreciable power to Earth's dynamo persisted through at least 14 Myr during the late Mesoproterozoic to maintain a strong surface field (Fig. \ref{fig:PINT_compilation}). In addition to these high geomagnetic fields recorded by Midcontinent Rift rocks, the ca. 755 Ma Mundine Well dikes \cite{Lloyd2021b} also require a stronger geomagnetic field in the Neoproterozoic than would be predicted by a progressive Proterozoic decline (Fig. \ref{fig:PINT_compilation}).

The hypothesis that a weak Ediacaran geomagnetic field is a telltale sign of the lack of an inner core with core nucleation following shortly thereafter may predict that it is the most significant weak to strong field transition in the paleointensity record. However, Fig. \ref{fig:PINT_compilation} shows that transitions from low to high field intensities occurred before, during, and after the Ediacaran. In the Ediacaran record developed to date, there is a two-fold increase in Earth's virtual dipole moment when comparing estimates from the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} to those from ca. 550 Ma volcanics of the Skinner Cove Formation \cite{Thallner2021a} (Fig. \ref{fig:PINT_compilation}). In the late Mesoproterozoic, there is at least a six-fold increase within a period of $\sim$35 Myr from a low average virtual dipole moment of $\sim$13 ZAm$^2$ recorded by the ca. 1140 Ma Abitibi dikes which yielded straight Arai plots \cite{Macouin2003a}, to a high moment of $\sim$70 ZAm$^2$ recorded by the ca. 1106 Ma Osler Volcanics, with even stronger values from ca. 1092 Ma by Beaver River anorthosite xenoliths that record virtual dipole moments up to $\sim$140 ZAm$^2$. While the ca. 1140 Abitibi dikes paleointensity estimates do not go to values as low as the ca. 565 Ma Sept-$\hat{I}$les intrusions, this virtual dipole moment increase in the Mesoproterozoic from the ca. 1140 Ma data to the ca. 1100 Ma data is the largest yet documented in the Precambrian on a 10s of millions of years timescale (Fig. \ref{fig:PINT_compilation}). The tempo and scale of this field intensity transition could match with model-based predictions associated with the onset of inner core nucleation \cite{Davies2021a}. This ca. 1.1 Ga timing would be broadly consistent with the ca. 1.3 Ga onset proposed by \citealp{Biggin2015a} albeit later given the exclusion of previous overestimated paleointensity values from the ca. 1.3 Ga Gardar basalts that are superseded by data in \cite{Kodama2019}. However, a model prediction of sustained strong field values following inner core nucleation is challenged by data from the ca. 1070 Ma Bangemall Sills which include a sill with a low virtual dipole moment of $\sim$6.4 ZAm$^2$ (Fig. \ref{fig:PINT_compilation};  \citealp{Lloyd2021b}). Following the Ediacaran, there are also low paleointensity estimates from Devonian rocks such as ca. 370 Ma dikes and lavas of the Siberian Viluy Traps that give virtual dipole moment estimates of 4.3 to 14.9 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2019a}). These low values as well as data from the ca. 414 Ma Strathmore lava flows \cite{Hawkins2021a}, the ca. 410-380 Ma lava flows of Siberia and the Kola Peninsula \cite{Shcherbakova2017a}, the ca. 408–393 Ma Buribay volcanics \cite{Shcherbakova2021a} and the ca. 332 Ma Kinghorn volcanics \cite{Hawkins2021a} has led to the proposal of this interval as the ``Mid-Paleozoic Dipole Low'' \cite{Hawkins2021a}. This ``Mid-Paleozoic Dipole Low'' is followed by high paleointensity values such that there is a six-fold increase in the database from virtual dipole moment of $\sim$16 ZAm$^2$ at ca. 332 Ma to $\sim$99 ZAm$^2$ ca. 308 Ma in the late Carboniferous (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2021a}).
